---
title: "St Croix Sed and Coral Analysis"
author: "Amanda Bayless"
date: "April 5, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(vegan) #Library required for PCA and RDA
library(dunn.test)
source("./utility_functions.R") #script file for PCA cleanplot

#Sediment Data
explanatory <- read.csv("../Data/Sediment Exp Matrix.csv")
community <- read.csv("../Data/Sediment Comm Matrix.csv")

#Coral Data
coral_explanatory <- read.csv("../Data/Coral Explanatory Variables.csv")
coral_community <- read.csv("../Data/Coral ICP Community Matrix.csv")
coral_community

Ca_ratio <- read.csv("../Data/Coral Community Matrix Ratios.csv") #This data is a ratio of all elements to Ca since the calcium looked variable in each site and was used for further hypothesis testing
```

#Create community and explanatory matrices for sediment analysis

```{r}
comm <- community[ , 3:ncol(community)] 
#Takes the community data set but only uses 3rd through last column

row.names(comm) <- comm[ , 1]
comm_sed <- comm[ , -1]
#Drop first column and make it a row name for community data

row.names(explanatory) <- explanatory[ , 1]
exp_sed <- explanatory[ , -1] 
#Drop first column and make sample.ID the row names for explanatory data (Don't want rows as explanatory variable!)

head(exp_sed) 
head(comm_sed)
#Checking successful move of first column to row for both community and explanatory matrix

all.equal(rownames(comm_sed), rownames(exp_sed)) 
#Checking row names and column names match up in size
```

#Indirect/Unconstrained Ordination on Sediment
Trying to understand key elements w/o knowing everything about the sediment data

```{r}
comm_pca <- rda(comm_sed) 
summary(comm_pca)
#Running PCA on community matrix; shows eigenvalues for each axis (Large eigenvalues correspond to large variances)
```

```{r}
sum(comm_pca$CA$eig) #sum of eigenvalues is equal to total variance
```

```{r}
round(comm_pca$CA$eig / comm_pca$tot.chi, 2)
#Ratio of eigenvalue to total variance shows amount of variance explained by each PCA axis
#Can see PC1 captures almost all variance
```

```{r}
plot(comm_pca,display=c('sp'))
#Elements seem to be oriented according to the 1 axis that explains almost all variance but still not revealing information about elements to eliminate. Iron and aluminum have greater abundance in nature than most other elements so it is obviously further along PC1 axis. 
```

```{r}
cleanplot.pca(comm_pca)
#You can see the clean plot a little easier by expanding this window
#The visible elements that stand out are the elements that should be focused on in sediment for examining significant differences between sites and between regions (BUIS and SARI)
```

#Constrained RDA with Site as predictor variable 
```{r}
rda_sed = rda(comm_sed ~ Site + Lat + Long, data = exp_sed)
# vegan requires that we write out each term if we are not going to convert the factor to a dummy matrix
# Will add grain size analysis as another predictor variable when data is available

rda_sed
# says "some constraints were aliased because they were collinear (redundant)" which means the lat and long were removed because site name is the same thing 

plot(rda_sed)
#This RDA is hard to read with each sample labeled on the plot
```

```{r}
plot(rda_sed, type='n', scaling=1)
orditorp(rda_sed, display='sp', cex=1, scaling=1, col='blue')
text(rda_sed, display='cn', col='red')
# This RDA plot is much cleaner, but still hard to view -> expanding window helps
# This plot does show that all of these elements are correlated with SARI sites and BUIS sites are not even visible. This indicates that all elements may be in higher concentrations at SARI sites (besides the one element hiding in the middle of the plot)
```

#Hypothesis Testing
The elements chosen for hypothesis testing are based on the results from the RDA as well as other elements that are of interest based on their role in sediment

*Numbers shown for all elements are in parts per billion (ppb)

COPPER - toxic to reproduction 
```{r}
plot(Cu ~ Site, ylab = "Cu (ppb)", data = community)
# Examining box plots to see element variation between sites

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Cu ~ Group, data = community, ylab = "Cu (ppb)", xlab = "Region")
# Also can examine how elements vary by region (SARI vs BUIS)
```

Copper looks like it varies a lot by site and region!

BARIUM - indication of freshwater (but mostly used in water column and in coral skeleton, not so much in sed)
```{r}
plot(Ba ~ Site, data = community, ylab = "Ba (ppb)")

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Ba ~ Group, data = community, ylab = "Ba (ppb)", xlab = "Region")
```


ZINC - Toxic metal
```{r}
plot(Zn ~ Site, data = community, ylab = "Zn (ppb)")

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Zn ~ Group, data = community, ylab = "Zn (ppb)", xlab = "Region")
```

VANADIUM - Can be toxic; mainly assoicated with oil refineries and mining
```{r}
plot(V ~ Site, data = community, ylab = "V (ppb)")

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(V ~ Group, data = community, ylab = "V (ppb)", xlab = "Region")
```

CHROMIUM - from industrial pollution (mainly toxic in hexavalent form)
```{r}
plot(Cr ~ Site, data = community, ylab = "Cr (ppb)")

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Cr ~ Group, data = community, ylab = "Cr (ppb)", xlab = "Region")
```

Large differences between sites and regions for Cu, Ba, V, Zn and Cr

#Test if these differences are significant with ANOVA

COPPER 
```{r}
Cu_aov1 <- lm(Cu ~ Site, data = community)
summary(Cu_aov1)

plot(Cu_aov1) # Data is not normally distributed and does not meet the assumptions of ANOVA. Cu is log transformed but data still does not look normally distributed, so we must perform a non-parametric test
shapiro.test(community$Cu) # p value way less than 0.05, so data is non-normal

dunn.test(community$Cu, community$Site, method = "bonferroni")

kruskal.test(Cu ~ Group, data = community)
```
SARI 8, 9, 11 are significantly different than BUIS 1 using Dunn's Test

By separating SARI and BUIS in 'Group', we can see that Cu concentrations inside Salt River Bay (SARI) are significantly different from Cu Concentrations at Buck Island sites (BUIS)

ZINC
```{r}
Zn_aov <- lm(Zn ~ Site, data = community)
summary(Zn_aov)

plot(Zn_aov) #Data not normally distributed. Can't log transform Zn and sqrt doesn't help

shapiro.test(community$Zn)

dunn.test(community$Zn, community$Site, method = "bonferroni")

kruskal.test(Zn ~ Group, data = community)
```
SARI 8 and 9 sig diff than BUIS 1; SARI 9 sig diff than BUIS 4

Zinc is also significantly different between BUIS and SARI and significantly different among sites

BARIUM
```{r}
Ba_aov <- lm(Ba ~ Site, data = community)
summary(Ba_aov)

plot(Ba_aov) #Data not normally distributed

shapiro.test(community$Ba)

dunn.test(community$Ba, community$Site, method = "bonferroni")

kruskal.test(Ba ~ Group, data = community)
```

SARI 8, 9 significantly different from BUIS 1 and BUIS 4

#Coral Analysis
All code below examines elements in coral skeleton

#Create community and explanatory matrices for coral
 
```{r}
coral_comm_short <- coral_community[ , 3:ncol(coral_community)] 
#takes the community data set but only uses 3rd through last column

coral_exp_short <- coral_explanatory[ , 3:ncol(coral_explanatory)]

row.names(coral_comm_short) <- coral_comm_short[ , 1] 
cor_comm <- coral_comm_short[ , -1] 
# drop first column and make it a row name

row.names(coral_exp_short) <- coral_exp_short[ , 1]
cor_exp <- coral_exp_short[ , -1] 
# make sample.ID the row names for explanatory matrix

head(cor_exp)
head(cor_comm)

all.equal(rownames(cor_exp), rownames(cor_comm)) 
#Checking row names and column names match up in size
```

#Indirect/Unconstrained Ordination on Coral
Trying to understand key elements to examine in coral skeleton

```{r}
coral_pca <- rda(cor_comm[ ,c(-3,-6,-13)]) #removed Calcium, Magnesium and Strontium in order to see other elements
coral_pca
#Running PCA on coral community matrix; shows eigenvalues for each axis (Large eigenvalues correspond to large variances) 
```

```{r}
round(coral_pca$CA$eig / coral_pca$tot.chi, 2)
```
May need to transform data or remove elements that dominate in coral skeleton (Ca, Sr, Mg) -> Just got data together yesterday, so still working on this issue and both plots below shows why this is necessary!

```{r}
plot(coral_pca,display=c('sp')) #Not a whole lot to take away from the PCA plot. 
```


```{r, warning = FALSE}
cleanplot.pca(coral_pca)
```

#Constrained RDA

This RDA examines elements as explanatory variables for colony length (cm) and percent mortality of the in each biopsy

Colony length and percent mortality were recorded right before taking biopsies of coral from the wild

```{r}
cor_exp1 <- na.omit(cor_exp) 
#omitting values with NA 
cor_exp1
cor_comm1 <- cor_comm[-c(55:64), ] 
cor_comm1
#Have to remove site BUIS 6 from community matrix because some information for the explanatory variables is not available and matrices must be equal in size in order to run RDA
```

```{r}
coral_rda = rda(cor_exp1$Perc.Mortality ~ Cu + Pb + Zn + Fe + Al + P + Ni + Co + Ba + Cd, data = cor_comm1) 
coral_rda
```

```{r, warning = FALSE}

plot(coral_rda)
```
Since unconstrained  variance is much greater then constrained variance, should use caution when interpreting results

We can see which colonies are correlated with which elements based on mortality. Row 1, 17 and 54 are correlated with higher Cu values (I can't get the names to show up) and these rows (colonies) had the greatest observed mortality in the data set. 

The same analysis will be run for the reproductive data and the element concentrations in the coral. 

Hypothesis testing still necessary...

#Hypothesis Testing with ANOVA 

First, graphing elements to see differences between sites

```{r}
plot(P ~ Site, xlab = "Site", ylab = "Phosphorus (ppb)", data = coral_community)

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(P ~ Group, xlab = "Population", ylab = "Phosphorus (ppb)", data = coral_community)

plot(Ba ~ Site, xlab = "Site", ylab = "Barium (ppb)", data = coral_community)

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Ba ~ Group, xlab = "Population", ylab = "Barium (ppb)", data = coral_community) 

plot(Zn ~ Site, xlab = "Site", ylab = "Zinc (ppb)", data = coral_community)

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Zn ~ Group, xlab = "Population", ylab = "Zinc (ppb)", data = coral_community) 


plot(Pb ~ Site, data = coral_community) 

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Pb ~ Group, xlab = "Population", ylab = "Lead (ppb)",data = coral_community)


plot(B ~ Site, data = coral_community)

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(B ~ Group, xlab = "Population", ylab = "Boron (ppb)", data = coral_community)

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Fe ~ Site, xlab = "Site", ylab = "Iron (ppb)", data = coral_community)

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Fe ~ Group, xlab = "Population", ylab = "Iron (ppb)", data = coral_community)

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Cu ~ Site, xlab = "Site", ylab = "Copper (ppb)", data = coral_community)

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Cu ~ Group, xlab = "Population", ylab = "Copper (ppb)", data = coral_community)

#Phosphorus, iron, barium, zinc, lead, and boron all appear to vary by site and region
```

```{r}
#The rest of the plots below did not show much difference between sites or regions

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Cd ~ Site, data = coral_community)

plot(Cd ~ Group, data = coral_community)

plot(Al ~ Site, data = coral_community)

plot(Al ~ Group, data = coral_community)

plot(Li ~ Site, data = coral_community)

plot(Li ~ Group, data = coral_community)

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Cu ~ Site, data = coral_community)

plot(Cu ~ Group, data = coral_community)

plot(Co ~ Site, data = coral_community)

plot(Ni ~ Site, data = coral_community)
```


Below are plots of the same elements in coral skeletons ratioed to calcium. The plots above of elements that aren't ratioed to Ca were kept for comparison to show committee. 

```{r}
plot(P ~ Site, xlab = "Population" , ylab = "P (ppb)", data = Ca_ratio)  
plot(P ~ Group, data = Ca_ratio) #Test significance

plot(Ba ~ Site, data = Ca_ratio) 
plot(Ba ~ Group, data = Ca_ratio) #Test significance

plot(Zn ~ Site, data = Ca_ratio)
plot(Zn ~ Group, data = Ca_ratio) #Test significance

plot(Pb ~ Site, data = Ca_ratio) 
plot(Pb ~ Group, data = Ca_ratio) #Test significance

plot(B ~ Site, data = Ca_ratio)
plot(B ~ Group, data = Ca_ratio) #Test significance

plot(Cd ~ Site, data = Ca_ratio)

plot(Fe ~ Site, data = Ca_ratio)

plot(Al ~ Site, data = Ca_ratio)

plot(Li ~ Site, data = Ca_ratio)

plot(Co ~ Site, data = Ca_ratio)

plot(Cu ~ Site, data = Ca_ratio)

plot(Ca ~ Site, data = Ca_ratio)

plot(Mg ~ Group, data = Ca_ratio)
```

COPPER
```{r}
Cu_coral <- lm(Cu ~ Site, data = Ca_ratio)
summary(Cu_coral)

plot(Cu_coral)

shapiro.test(Ca_ratio$Cu) # not normally distributed

dunn.test(Ca_ratio$Cu, Ca_ratio$Site, method = "bonferroni")
#Cu in coral not significantly diff btwn sites

kruskal.test(Cu ~ Group, data = Ca_ratio)
#Cu in coral not significantly different between SARI and BUIS as a whole
```
Cr and V are also not significantly different in coral at SARI vs BUIS

ZINC
```{r}
Zn_coral <- lm(sqrt(Zn) ~ Site, data = Ca_ratio)
summary(Zn_coral)

plot(Zn_coral)

shapiro.test(Ca_ratio$Zn) #not normally distributed

dunn.test(Ca_ratio$Zn, Ca_ratio$Site, method = "bonferroni")

kruskal.test(Zn ~ Group, data = Ca_ratio)

#Zn is significantly greater at SARI  
```
SARI 2 sig diff than BUIS 6; SARI 3 sig diff than BUIS 2 and BUIS 6 

Phosphorus - Indicates nutrient loading 
```{r}
P_coral <- lm(P ~ Site, data = Ca_ratio)
summary(P_coral)

plot(P_coral)

shapiro.test(Ca_ratio$P) #close to being normal, may be able to log transform? but can't seem to log transform  in shapiro test function 

dunn.test(Ca_ratio$P, Ca_ratio$Site, method = "bonferroni")

P_coral2 <- lm(P ~ Group, data = Ca_ratio)
summary(P_coral2)

plot(P_coral2)
#P significantly greater at BUIS
```

Boron - pH proxy
```{r}
B_coral <- lm(B ~ Site, data = Ca_ratio)
summary(B_coral)

plot(B_coral)

shapiro.test(Ca_ratio$B) #Data normally distributed

B_coral2 <- lm(B ~ Group, data = Ca_ratio)
summary(B_coral2)

plot(B_coral2)
#B significantly greater at BUIS 
```

BARIUM - Salinity/Freshwater proxy 
```{r}
Ba_coral <- lm(log(Ba) ~ Site, data = Ca_ratio)
summary(Ba_coral)

plot(Ba_coral)

shapiro.test(Ca_ratio$Ba) #Data not normally distributed

dunn.test(Ca_ratio$Ba, Ca_ratio$Site, method = "bonferroni")

kruskal.test(Ba ~ Group, data = Ca_ratio)

Ba_coral2 <- lm(Ba ~ Group, data = Ca_ratio)
summary(Ba_coral2)
#Ba significantly greater at BUIS
```
SARI 1 sig diff from BUIS 1 and BUIS 2; SARI 2 sig diff from BUIS 2

LEAD - Toxic metal 
```{r}
Pb_coral <- lm(Pb ~ Site, data = Ca_ratio)
summary(Pb_coral)

shapiro.test(Ca_ratio$Pb) #Data normally distributed

pairwise.t.test(Ca_ratio$Pb, Ca_ratio$Site) #sig differences btwn sites; double check all assumptions

dunn.test(Ca_ratio$Pb, Ca_ratio$Site, method = "bonferroni") 

kruskal.test(Pb ~ Group, data = Ca_ratio)

Pb_coral2 <- lm(Pb ~ Group, data = Ca_ratio)
summary(Pb_coral2)
#Pb significantly greater at SARI
```

ALUMINUM AND IRON - Both element indicate turbidity if taken up in coral skeleton because they are the most abundant elements in muddy sediment 
```{r}
Al_coral <- lm(Al ~ Site, data = Ca_ratio)
summary(Al_coral)

shapiro.test(Ca_ratio$Al) #Data normally distributed

dunn.test(Ca_ratio$Al, Ca_ratio$Site, method = "bonferroni") #no sig diff btwn sites 

kruskal.test(Al ~ Group, data = Ca_ratio)
```
No sig differences for Al

```{r}
Fe_coral <- lm(Fe ~ Site, data = Ca_ratio)
summary(Fe_coral)

shapiro.test(Ca_ratio$Fe) #Data not normally distributed

dunn.test(Ca_ratio$Fe, Ca_ratio$Site, method = "bonferroni") #no sig diff btwn sites 

kruskal.test(Fe ~ Group, data = Ca_ratio) #significantly different when running kruskal test

Fe_coral2 <- lm(Fe ~ Group, data = Ca_ratio) #not significantly diff when running parametric 

summary(Fe_coral2)
```

Examining percent mortality (explanatory variable) as a function of site

```{r}
plot(Perc.Mortality ~ Site, data = coral_explanatory) 
#Mortality appears to vary by site

Mort_aov <- lm(Perc.Mortality ~ Site, data = coral_explanatory)
summary(Mort_aov)
#Mortality is significantly different between certain sites 

plot(Mort_aov)

shapiro.test(coral_explanatory$Perc.Mortality) #Data not normally distributed

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Perc.Mortality ~ Site, ylab = "Percent Mortality", data = coral_explanatory)
dunn.test(coral_explanatory$Perc.Mortality, coral_explanatory$Site, method = "bonferroni") 

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Perc.Mortality ~ Group, ylab = "Percent Mortality", xlab = "Population", data = coral_explanatory)
kruskal.test(Perc.Mortality ~ Group, data = coral_explanatory)

```
SARI 2 and BUIS 4 are sig diff than BUIS 1
BUIS 4 and SARI 2 all have significantly greater mortality than BUIS 1

```{r}
plot(Colony.length ~ Site, data = coral_explanatory)
# Colony length appears to vary slightly

Length_aov <- lm(Colony.length ~ Site, data = coral_explanatory)
summary(Length_aov)

plot(Length_aov)

shapiro.test(coral_explanatory$Colony.length) #Data not normally distributed

dunn.test(coral_explanatory$Colony.length, coral_explanatory$Site, method = "bonferroni")

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Colony.length ~ Group, xlab = "Population", ylab = "Colony Length (cm)", data = coral_explanatory)

par(cex.lab = 1.5)
par(cex.axis = 1.5)
plot(Colony.length ~ Site, ylab = "Colony Length (cm)", data = coral_explanatory)

kruskal.test(Colony.length ~ Group, data = coral_explanatory)
```

SARI 1 is significantly diff than BUIS 1 and BUIS 2
SARI 1 is significantly smaller in length than BUIS 1 (and BUIS 2?)

Sediment Notes:

U significantly different at individual sites but not between BUIS and SARI 
Mg, Nd, Ag not significantly different
Rb diff at SARI 1 but not between BUIS and SARI 
Sr diff at a couple sites, not  btwn BUIS and SARI 
Li diff at a few sites, not btwn BUIS and SARI 

